Estimating duration using survival analysis

### 2024-03-29 cis-perfect-testing.tex
new likelihood in main text

In this section, I develop a computationally feasible Bayesian statistical model accounting for double interval censoring and undetected episodes.
In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
Then, in \cref{perf-test:sec:episode-types}, I adapt the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
The number of such episodes in each class has a multinomial distribution conditional on the latent total number of infections (including undetected infections).

In \cref{perf-test:sec:posterior-density}, I outline the form of the posterior density.
The following subsections then derive its various components.
Unlike in \textcite{heiseyModelling}, I must consider the individual's probability of undetected episodes separately because they have differing visit times.
In addition, I include the information from the individuals without a detected episode.

@@

\subsection{Data and notation} \label{perf-test:sec:data-notation}

@@

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
These are all the times an infection episode can be detected at, plus one prior test.
For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

For a detected episode $j$, five pieces of data are observed (see \cref{E-biology-data:sec:cis-episodes} for how these are constructed).
The individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
Over the short calendar period considered, each individual has zero or one detected episodes.
Let $i \in \set{D}$ indicate individual $i$ has a detected episode and denote $j(i)$ as the detected episode in individual $i$; $j(i)$ is only defined for $i \in \set{D}$.

In addition, the individuals in the cohort experienced a total of $\nnodet$ undetected infection episodes.
$\nnodet$ is a latent quantity.
The total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.

@@

\subsection{Types of episodes} \label{perf-test:sec:episode-types}

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$), shown in \cref{perf-test:fig:partitionSpace}.
Each episode, whether detected or not, must fall into one of these three classes with a probability that depends on the parameters of the survival function, $\vec{\theta}$, specified in \cref{perf-test:sec:parameters-priors}.
The regions are defined for each individual $i$.

\begin{itemize}
\item
  The region of admissible episodes for individual $i$, $\alpha_i$.
  An admissible episode has an infection and recovery time consistent with the data observed.
  Let $n_{ia}$ be the number of admissible episodes for $i$.
  $\alpha_i$ is empty if $i$ had no detected episodes (\ie $i \notin \set{D}$); in this case the observation is $n_{ia} = 0$.
  Otherwise, if $i \in \set{D}$, then $n_{ia} =1$.
  The admissible region is the intersection of the intervals $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$ and $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
  This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
  The shape of this region accounts for the doubly interval censored nature of the data.
  Let $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \alpha_i$ given that $j$ occurs in individual $i$.
  \label{perf-test:def:admissible}
\item
  The region of undetected episode for individual $i$, $\Omega_i$.
  Let $n_{iu}$ be the (latent) number of undetected episodes for $i$.
  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, indicating combinations of episode beginning and end times which do not span a testing time plus episodes beginning before the individual's enrolment.
  Let $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \Omega_i$ given that $j$ occurs in individual $i$.
\item
  The region of inadmissible episodes, $\beta_i$ for individual $i$.
  An inadmissible episode is one which would have been detected had it occurred, but the data rule out having occurred.
  Let $n_{i\bar{a}}$ be the number of inadmissible episodes for $i$, which is observed as $n_{i\bar{a}} = 0$.
  These are all remaining (a priori possible) episodes not in the previous sets.
  Let $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \beta_i$ given that $j$ occurs in individual $i$.
\end{itemize}

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and undetected regions]{%
  The regions for individual $i$.
  $i$'s first test was at time 0 and negative, with subsequent negative tests at times 7, 14, and 56, and positive tests at times 21 and 28.
  Inadmissible regions are unshaded.
  The grey (impossible) region covers the cases in which recovery occurs prior to infection.
  \label{perf-test:fig:partitionSpace}
}
\end{figure}

These three regions span all possible episodes and are mutually exclusive.
Hence the following relations hold:
\begin{align}
\ntot
&= \ndet + \nnodet + \sum_{i=1}^{\Ncis} n_{i\bar{a}} \\
&= \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0
\end{align}
and, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.

I will use these regions to form the data likelihood, conditional on $\ntot$.
The definition of the admissible region relies on the test results.
However the likelihood should be conditional only on $\vec{\theta}$, $\ntot$, and the test schedules.
Therefore, this is not strictly correct.
This issue can be rectified by considering the inadmissible region as many subregions, which leads to the same result (sketched in \cref{perf-test:sec:alt-likelihood}).
A full formalisation of the derivation in \cref{perf-test:sec:alt-likelihood} would require much more complex notation for the same result; therefore, I prefer the presentation to follow.

% Any episode $j$ independently belongs to one of these classes.
The probabilities $p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$ are calculated conditional on the episode occurring in individual $i$.
I assume the individual in which an episode occurs is iid uniform across the cohort, that is $\prob(i(j) = i) = 1/\Ncis$.
Therefore, the conditional probabilities are multiplied by $1/\Ncis$ to form the probabilities .
For example, $\prob((B_j, E_j) \in \alpha_i, i(j) = i \mid \vec\theta) = p_{ia} / \Ncis$.

These probabilities apply to each episode independently.
Therefore:
\begin{align}
  \begin{bmatrix}
    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
  \end{bmatrix}
  \mid \ntot, \vec{\theta}
  \dist
  \MN\left(
    \ntot,
    \frac{1}{\Ncis}
    \begin{bmatrix}
      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
    \end{bmatrix}
  \right)
  \label{perf-test:eq:multinomial}
\end{align}

The undetected regions cannot be distinguished.
Therefore, I combine these regions into a single cell within the multinomial distribution.
The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability parameter $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.

\subsection{Posterior density} \label{perf-test:sec:posterior-density}

Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
This represents all the observations used in \cref{perf-test:eq:multinomial}.
Therefore, the posterior density we are interested in is $p(\vec{\theta} \mid \na)$.
It can be written as follows.
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) p(\na \mid \vec{\theta}) \\
&= p(\vec\theta) \sum_{\ntot= \ndet}^{\infty} p(\ntot, \na \mid \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta})
\label{perf-test:eq:posterior1}
\end{align}
where the last line augments the likelihood with the total number of episodes, $\ntot$.

First, I derive the form of $p(\na \mid \ntot, \vec{\theta})$.
From \cref{perf-test:sec:episode-types}, the counts $\na$ and $\nnodet$'s joint distribution, conditional on $\ntot$, is multinomial.
All $n_{i\bar{a}} = 0$, therefore these classes can be ignored; counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\Ncis} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
Hence:
\begin{align}
p(\na \mid \ntot, \vec{\theta})
&= p(\nnodet  = \ntot - \na, \na\mid \ntot, \vec{\theta}) \\
&= \frac{\ntot!}{(\ntot - \ndet)! \prod_{i=1}^{\Ncis} n_{ia}!}  \pnodet^{\ntot - \ndet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
\end{align}
the final line follows because $n_{ia} = 1$ for $i \in \set{D}$ and $n_{ia} = 0$ otherwise.

Substituting \cref{perf-test:eq:augmented-likelihood} into \cref{perf-test:eq:posterior1}:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
&= p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:posterior2}.
\intertext{For convenience, define the summation term as:}
\eta &= 
\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
\label{perf-test:eq:eta}
\end{align}
The rest of this section derives expressions for each of $p_{ia}$, $p_{u}$ and $\eta$.

@@

Substituting this into \cref{perf-test:eq:posterior2} gives:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior}.
\end{align}

@@

\subsection{Admissible probabilities}

Next, I derive $p_{ia}$.
From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.

By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
\begin{align}
p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
p_{ia}
=& \prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \right) \\
=& \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

\subsection{Undetected probability} \label{perf-test:sec:prob-undetected}

The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
From \cref{perf-test:sec:episode-types}:
\begin{align}
  1 - p_u
  &= 1 - \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu} \\
  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - p_{iu}).
  \label{perf-test:eq:pu}
\end{align}

Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of detecting an episode.

An episode $j$ in individual $i(j)$ is detected if and only if all the following conditions are met.
\begin{enumerate}
    \item $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$; this condition is equivalent to having at least one positive test for the episode.
    \item $\exists t \in \sched_{i(j)}$ such that $t < b_j$.
      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded; thereby adjusting for those who enrolled during the period being less likely to have a detected infection.
      For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
    \item $\exists t \in \sched_{i(j)}$ such that $t \geq b_j$.
      This means that the test is detected before $T$.
    \item Individual $i(j)$ is tested at a time $t > e_j$, upper bounding the end of the episode.
      Note that this could be after $T$, and hence it is not necessarily the case that $t \in \sched_{i(j)}$.
      For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
      % For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
\end{enumerate}

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\label{perf-test:eq:tau-def}
\end{align}
defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
The fourth condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
&= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \tau_{\sched_{i(j)}}(b_j) < \infty, B_j > \min \sched_{j}) \\
&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1, \tau_{\sched_{i(j)}}(b_j) < \infty \mid B_j = b) \prob(B_j = b)\\
&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
\label{perf-test:eq:piu}
\end{align}
defining $S_{\vec{\theta}}(\infty) = 0$.

Substituting into \cref{perf-test:eq:pu}:
\begin{align}
1 - p_u
& \propto \sum_{i=1}^{\Ncis} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
\end{align}

>>>>

In this section, I develop a computationally feasible Bayesian statistical model accounting for double interval censoring and undetected episodes.
In \cref{perf-test:sec:data}, I introduce the data, defining which quantities are observed and which are latent.
Then, in \cref{perf-test:sec:episode-types}, I build a framework, using the concept of \textcite{heiseyModelling}, which classifies possible episode beginning and end time into \emph{admissible} regions, each of which match a possible pattern of the doubly interval censored pattern of the observations; and an \emph{undetected} region, which lead to the episode being undetected.
The number of such episodes in each region has a multinomial distribution conditional on the latent total number of infections (including undetected infections).

In \cref{perf-test:sec:posterior}, I outline the form of the posterior density.
The following subsections then derive its various components.
Unlike in \textcite{heiseyModelling}, I must consider the individual's probability of undetected episodes separately because they have differing visit times.
In addition, I include the information from the individuals without a detected episode.

@@

\subsection{Data} \label{perf-test:sec:data}

@@

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise.
Note that $\sched_i$ includes any tests that occur after $T$.
For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

For any episode $j$, define $O(j)$ as the observations of that episode.
Test results are deterministic based on the time of the test and the individual's infection status.
Therefore, $O(j)$ is fully determined by the triplet $(B_j, E_j, i(j))$ consisting of $j$'s beginning time, end time, and the individual in which the episode occurs respectively.

For a detected episode, we have $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$.
These are: the individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
How the episode data are constructed is detailed in \cref{E-biology-data:sec:cis-episodes}.

For an undetected episode $j$, there are no observations; in which case, define $O(j) = \emptyset$.
Let $\nnodet$ denote the (latent) number of undetected episodes experienced across all $\Ncis$ individuals.
Therefore, the total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.

\subsection{Posterior density} \label{perf-test:sec:posterior}

@@

Define an integer $N_E$ and $\set{E} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_E} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{E}$ if and only if $j$ is a detected infection.
Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$ be an arbitrary vector of possible observations, visualised in \cref{perf-test:fig:partitionSpace}; the conditions for $\vec{\nu}_k \in \set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.

The target of inference is $\vec{\theta}$, the parameters of the survival function $S_{\vec{\theta}}(t)$.
The contents of $\vec{\theta}$ are discussed in \cref{perf-test:sec:parameters-priors}.

Let $n_k$ denote the number of observations of $\vec{\nu}_k$; $\nnodet$ denote the latent number of undetected episodes; and $\vec{n} = [n_1, \dots, n_{N_D}, \nnodet]^T$.
Hence, $\ntot = \sum_{i=1}^{N_D} n_i + \nnodet$.
For $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$; $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$; and $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
Then:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and undetected regions]{%
  The regions for individual $i$.
  $i$'s first test was at time 0 and negative, with subsequent negative tests at times 7, 14, and 56, and positive tests at times 21 and 28.
  Inadmissible regions are unshaded.
  The grey (impossible) region covers the cases in which recovery occurs prior to infection.
  \label{perf-test:fig:partitionSpace}
}
\end{figure}

In the CIS data, each $n_k$ ($k \neq u$) is observed as either 0 or 1.
Define $D = \{ k \ssep n_k = 1 \}$.
Furthermore, note that the support of the multinomial distribution requires that $\nnodet = \ntot - \ndet$.
Let $\na = [n_{1}, \dots, n_{N_E}]^T$ be the observed portion of $\vec{n}$.
Then \cref{perf-test:eq:multinomial-ll} simplifies to:
\begin{align}
  p(\vec{n} \mid \ntot, \vec{\theta})
  &= p(\na \mid \ntot, \vec{\theta}) \\
  &= \frac{\ntot!}{(\ntot - \ndet)!} p_u^{\ntot-\ndet} \prod_{k \in \set{D}} p_k.
  \label{perf-test:eq:multinomial}
\end{align}

The relevant information from the CIS data is fully contained in the vector $\na$.
Therefore, the posterior of interest is:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) p(\na \mid \vec{\theta}) \\
&= p(\vec\theta) \sum_{\ntot= \ndet}^{\infty} p(\ntot, \na \mid \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{k \in \set{D}} p_k &\text{by \cref{perf-test:eq:multinomial}} \\
&= p(\vec{\theta}) \left( \prod_{k \in \set{D}} p_k \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right)
\label{perf-test:eq:posterior1}
\intertext{For convenience, define the summation term as:}
\eta &= 
\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
\label{perf-test:eq:eta}
\end{align}
The rest of this section derives expressions for each of $p_{k}$, $p_{u}$ and $\eta$.

@@

Substituting this into \cref{perf-test:eq:posterior1} gives:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_k \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior}.
\end{align}

@@

\subsection{Deriving $p_k$}

Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ are the observations of episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$.

Decompose $p_k$ as $p_k = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$
where $p_{ik} = \prob(O(j) = \nu_k \mid i(j) = i_k, \vec{\theta})$.
% This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.

If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{e}_k]$.
Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
\begin{align}
p_{ik}
=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
=& \sum_{b = l_k^{(l)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

\subsection{Deriving $1 - p_u$} \label{perf-test:sec:prob-undetected}

The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
Recall the definition $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Therefore:
\begin{align}
  1 - p_u
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset, i(j) = i \mid \vec{\theta}) \\
  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) \\
  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}))
  \label{perf-test:eq:pu}
\end{align}
assuming $P(i(j) = i \mid \vec{\theta})$ as before.
Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta})$.
Therefore, the crucial component is $1 - p_{iu}$.
This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of the episode being detected.

An episode $j$ in individual $i(j)$ is detected if and only if all the following conditions are met.
\begin{enumerate}
    \item $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$; this condition is equivalent to having at least one positive test for the episode.
    \item $b_j > \min \sched_{i(j)}$.
      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded; thereby adjusting for those who enrolled during the period being less likely to have a detected infection.
      For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
    \item $b_j \leq T_{i(j)}$ where $T_{i(j)} = \max \{ t \in \sched_{i(j)} \ssep t \leq T \}$ is the last time that $i(j)$ is tested in the period, meaning that the test is detected within the period.
    \item Individual $i(j)$ is tested at a time $t > e_j$, upper bounding the end of the episode.
      Note that this could be after $T$, and hence it is not necessarily the case that $t \in \sched_{i(j)}$.
      For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
      % For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
\end{enumerate}

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\label{perf-test:eq:tau-def}
\end{align}
% defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
% The fourth condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
% &= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \min \sched_{i(j)} < B_j \leq T_{i(j)}) \\
&= \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} \prob(D_j \geq \tau_{\sched_{i}}(b) + 1, \tau_{\sched_{i(j)}}(b_j) < \infty \mid B_j = b) \prob(B_j = b)\\
&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
\label{perf-test:eq:piu}
\end{align}

Substituting into \cref{perf-test:eq:pu}:
\begin{align}
1 - p_u
& \propto \sum_{i=1}^{\Ncis} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
\end{align}
